---
title: "MovieLens Analysis"
author: "Derrick"
date: "2023-12-12"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MovieLens Data Analysis

This dataset describes 5-star rating and free-text tagging activity from [MovieLens](http://movielens.org), a movie recommendation service. It contains 100836 ratings and 3683 tag applications across 9742 movies. These data were created by 610 users between March 29, 1996 and September 24, 2018. This dataset was generated on September 26, 2018.

The data are contained in the files links.csv, movies.csv, ratings.csv and tags.csv

```{r Loading the packages}
install.packages("tidyverse")
install.packages("wordcloud")
install.packages("wordcloud2")
install.packages("tm")
install.packages("slam")
library(dplyr)
library(pillar)
library(ggplot2)
library(tidyverse)
library(readr)
library(tidyr)
library(lubridate)
library(wordcloud)
library(wordcloud2)
library(tm)
library(slam)
```

```{r Importing the data}
ratings <- read_csv("ratings.csv")
movies <- read_csv("movies.csv")
links <- read_csv("links.csv")
tags <- read_csv("tags.csv")
```

```{r Exploring the Data}
str(links)
str(movies)
str(ratings)
str(tags)

# Since the links and the movies have same number of observations and variables they can be merged into one data
Mlinks <- merge(links, movies, by=c('movieId'))

# The ratings and the tags can also be merged into one data
Rtags <- bind_rows(ratings, tags)
```

### Cleaning the Rtags Dataset

```{r Handling missing values}
# Identify missing values in Rtags
missing_values <- colSums(is.na(Rtags))

# Replace missing values with a specific value (e.g., 0)
Rtags$rating[is.na(Rtags$rating)] <- 0

# Replace missing values with a specific value (e.g., no)thing
Rtags$tag[is.na(Rtags$tag)] <- "no tag"

# View columns with missing values
print(missing_values[missing_values > 0])
```

### Cleaning the Mlinks Dataset

```{r feature engineering}
# Cleaning the genres column
Mlinks$SingleGenre <- sapply(strsplit(as.character(Mlinks$genres), "|", fixed = TRUE), function(x) x[1])
```

```{r continuation}
# Use separate function to split the title column
Mlinks <- separate(Mlinks, title, into = c("Title", "Year"), sep = " \\(")

# Remove the closing parenthesis from the "Year" column
Mlinks$Year <- gsub("\\)", "", Mlinks$Year)

# Print the resulting dataframe
head(Mlinks)
```

```{r Handling missing values in Mlinks}
# Year the column to be modified
Mlinks$Year <- as.numeric(gsub("[^0-9]+", "0", Mlinks$Year))

# Fill in missing values with 0
Mlinks$Year[is.na(Mlinks$Year)] <- 0

# Fill in missing values with 0
Mlinks$tmdbId[is.na(Mlinks$tmdbId)] <- 0

# Identify missing values in Mlinks
missing_values <- colSums(is.na(Mlinks))
# View columns with missing values
print(missing_values[missing_values > 0])
```

```{r simple statistics}
summary(Mlinks)
summary(Rtags)
```

```{r Handling Duplicate Data}
# Identify and remove duplicate rows
Rtags <- distinct(Rtags)
Mlinks <- distinct(Mlinks)
```

```{r EDA}
# Univariate Analysis
# Visualize the distribution of numerical variables
# Example: Histogram for a numerical variable
ggplot(Rtags, aes(x = rating)) +
  geom_histogram(binwidth = 0.5, fill = "lightblue", color = "black") +
  labs(title = "Distribution of ratings")
```

```{r Violin plot}
ggplot(Rtags, aes(x = 1, y = rating)) +
  geom_violin(color="black",fill="lightblue") +
  labs(title = "Violin Plot to show rating distribution", y = "rating")
```

```{r Genre Distribution}

Count_movies <- Mlinks %>% 
                    group_by(SingleGenre) %>% 
                            dplyr::summarize(Total = n())

# Define color palette for genres
genre_colors <- c(
 "Western" = "darkblue",
 "War" = "red",
 "Thriller" = "green",
 "Sci-Fi" = "purple",
 "Romance" = "orange",
 "Mystery" = "yellow",
 "Musical" = "pink",
 "Horror" = "#c77dff",
 "Film-Noir" = "lightblue",
 "Fantasy" = "darkblue",
 "Drama" = "darkgreen",
 "Documentary" = "lightgreen",
 "Crime" = "#1d3557",
 "Comedy" = "black",
 "Children" = "#ff006e",
 "Animation" = "#8d99ae",
 "Adventure" = "#f2e8cf",
 "Action" = "brown",
 "(no genres listed)" = "#aacc00"
)

ggplot(Mlinks, aes(x = SingleGenre, fill = SingleGenre)) +
  geom_bar(width = 0.9) +  # Wider bars
  labs(title = "Distribution of Movies by Genre", x = "Genre", y = "Count") +
  coord_flip() +
  scale_fill_manual(values = genre_colors) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  ggtitle("Distribution of Movies by Genre") 
```

### A wordcloud to show tags

```{r word cloud}

tag_data <- tags$tag
tag_data <- as.data.frame(tag_data)

word_freq <- table(tag_data$tag)
word_freq <- as.data.frame(word_freq)

# Generate word cloud
cloud <- wordcloud(
  words = tag_data,
  min.freq = 2,
  scale = c(3, 0.5),
  colors = brewer.pal(8, "Dark2"),
  random.order = FALSE,
  max.words = 100,
  coords = NULL # Initially leave coords empty
)
```
